---
title: "Modelling Estuarine CHLa in the Upper Reaches of the Lower James"
subtitle: "James River Station JMS75 (Rice River Center), 2010-2023"
editor_options: 
  chunk_output_type: console
output: 
  html_document:
    toc: TRUE
    toc_depth: 4
    toc_float: true
---

```{r setup, include = FALSE, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, message = FALSE, warning = FALSE)

# Connect to 'functions.R' file so code can interact with custom functions. The file should be contained in the working directory, along with this .rmd file.
if (file.exists("functions.R")) {
  source("functions.R")
} else {
  stop("The 'functions.R' file is missing from the working directory. Call `getwd()` to check path of current working directory.")
}

# Install any missing packages
install_packages <- function(packages) {
  new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
  if(length(new_packages) > 0) {
    install.packages(new_packages, dependencies = TRUE)
  }
}

packages <- c("tidyverse", "ggplot2", "openxlsx", "kableExtra",  "zoo")
install_packages(packages)

library(ggplot2)
library(tidyverse)
library(kableExtra)

```

<br>

```{r `read data`}
# Read in Excel file and prepare data for analysis
excel_file <- "CHLa_data.xlsx"  # <--- change as needed depending on your Excel workbook file name

  
  # Define the two source dataframes
  # 1 df.james
  df.james <- openxlsx::read.xlsx(excel_file, sheet = 2)
        
      # Convert `date` column from Excel encoded date to a more legible date format. Otherwise date shows as numeric value, e.g. '44230'.
      df.james$date <- df.james$date * 86400      # 86400 = seconds in a day.
      df.james$date <- as.POSIXct(df.james$date, origin = "1899-12-30", tz = "UTC")
        
  # 2 df.discharge
  df.discharge <- openxlsx::read.xlsx(excel_file, sheet = 3)
  
  # Define column names 
    colnames(df.james)[3:5] <- c("SurfaceTemp", "CHLa", "DIN")
                       
    colnames(df.discharge) <- as.character(colnames(df.discharge))
    
    
  # Define and set dynamic variables for use in e.g, for loops.
    yr_range <-range(df.james$date)
    min_yr <- substr(yr_range[1], 1, 4)
    max_yr <- substr(yr_range[2], 1, 4)
     
    years_vec <- c(min_yr:max_yr)
    
                print(yr_range)
                print(min_yr)
                print(max_yr)
                print(years_vec)

```

```{r `preprocess data`}
  # Set and populate new fields for ln(CHLa),discharge, and day of year (1-365) 
  df.james <- 
    df.james %>% 
      mutate(ln_chla = log(df.james$CHLa),
             Discharge = as.numeric(NA),
             yday = yday(date) # `yday()` function returns day of year for a given date. )
             )
  
  # Add discharge values 
  for (i in years_vec) {
    ## subset DOYs by year and assign to variable
    DOY_vec <- df.james$yday[year(df.james$date) == i]
  
    ## Add discharge values to data frame
    df.james$Discharge[year(df.james$date) == i] <- df.discharge[DOY_vec, as.character(i)]
}
  
  
  # Create factor for differentiating data points before and after 2018
  df.james$period <- NA
  
    for (i in seq_along(df.james$date)) {
      df.james$period[i] <- ifelse(
        year(df.james$date[i]) %in% c(2010:2018), 
        "2010-2018", 
        paste0("2019-", as.character(max_yr)))
    }


```

## CHLa and DIN Time Series Plots

```{r `CHLa time series plot`, fig.width=10.5, fig.height=4.5, warning=FALSE, include = TRUE}

# Generate plot using ggplot package 
  suppressWarnings(
    ggplot(data = df.james) +
    geom_point(aes(x = date, y = CHLa,
                   color = CHLa) ) +
    geom_line(aes(x = date, y = CHLa,
                   color = CHLa) ) +
    geom_hline(yintercept = 40, linetype = "dashed", color = "gray") +
    labs(title = NULL, # or set title with quoted character string 
         y = paste0("Total CHLa (\u03BCg/L)"),
         x = "Date",
         ) +
    scale_y_continuous(breaks = seq(0, max(df.james$CHLa, na.rm = TRUE), by = 20)) +
    theme_classic() +
    theme(legend.position = "none",
                legend.text = element_text(size = 10),
                plot.title = element_text(size = 12),
                axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
                axis.text.y = element_text(size = 11),
                axis.line = element_line(color = "grey"),
                axis.title.x = element_text(size = 15, color = "grey20", margin = margin(t = 15)),
                axis.title.y = element_text(size = 15, color = "grey20", margin = margin(r = 15))) +
    scale_x_datetime(date_labels = "%Y", date_breaks = "1 years",
                     limits = c(as.POSIXct(paste0(min_yr, "-01-01")), as.POSIXct(paste0(max_yr,"-12-31"))
                                ) ) +
   scale_color_viridis_c(option = "D", direction = 1)  # Set color palette
   )

# Save plot to working directory
ggsave(filename = "output/CHLa_timeseries.jpeg", device = "jpeg", dpi = 300)

```

<br>

```{r `DIN time series plot`, fig.width=10.5, fig.height=4.5, include = TRUE, warning=FALSE}

  suppressWarnings(
    ggplot(data = df.james) +
    geom_point(aes(x = date, y = DIN,
                   color = DIN), alpha = .8 ) +
    geom_line(aes(x = date, y = DIN,
                   color = DIN) ) +
    labs(title = NULL, # or set title with quoted character string 
         y = "Total DIN (mg/L)",
         x = "Date",
         color = "Year") +
    scale_y_continuous(breaks = seq(0, max(df.james$DIN, na.rm = TRUE), by = .25)) +
    theme_classic() +
    theme(legend.position = "none",
                legend.text = element_text(size = 10),
                plot.title = element_text(size = 12),
                axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
                axis.text.y = element_text(size = 11),
                axis.line = element_line(color = "grey"),
                axis.title.x = element_text(size = 14, color = "grey20", margin = margin(t = 15)),
                axis.title.y = element_text(size = 14, color = "grey20", margin = margin(r = 15))) +
    scale_x_datetime(date_labels = "%Y", date_breaks = "1 years",
                     limits = c(as.POSIXct(paste0(min_yr, "-01-01")), as.POSIXct(paste0(max_yr,"-12-31"))
                                ) ) +
   scale_color_gradient(low = "#40798C", high = "grey20")  # manually set the two ends of the color ramp
  )

```

<br><br>

## Summary Statistics by Year

```{r `summary statistics/mean vals`}
 
  mean_vals <- data.frame(Year = years_vec)
  jul1_yday <- yday("2013-07-01")
  oct31_yday <- yday("2023-10-31")
  
  # Calculate mean discharge by year, Jul-Oct (all dates)  
    for (i in 2:ncol(df.discharge)) {
      mean_vals$Discharge[i-1] <- 
        round(mean(df.discharge[jul1_yday:oct31_yday, i], na.rm=T), 2)
    }
  
        
  # Calculate mean discharge by year based only on CHLa sampling days during Jul-Oct.
     df.samplingDays <- 
          data.frame(
            DOY = yday(df.james$date[yday(df.james$date) >= jul1_yday & 
                                          yday( df.james$date) < oct31_yday] ),
            year = year(df.james$date)[yday(df.james$date) >= jul1_yday & 
                                          yday( df.james$date) < oct31_yday]
          ) 
        
        for (i in (min(years_vec):max(years_vec)) ) {
              df.samplingDays$discharge[df.samplingDays$year == i] <-
                df.discharge[
                  df.samplingDays$DOY[df.samplingDays$year == i], as.character(i)]
          }
        
        for (i in 2:ncol(df.discharge)) {
              doy <- 
                df.samplingDays$DOY[df.samplingDays$year == 
                                      as.numeric(colnames(df.discharge)[i])]
              
              mean_vals$Discharge_sampling[i-1] <- 
                round(
                  mean(df.discharge[doy, i], na.rm = TRUE), 2)
  }

     
  # Calculate mean for variables with weekly observation data
  for (col in c("SurfaceTemp", "CHLa", "DIN", "SpCond") ) {
    means <- df.james %>%
        group_by(year = lubridate::year(date) ) %>%
        filter(month(date) %in% c(7:10) ) %>%
        summarize(mean_value = mean(!!sym(col), na.rm = TRUE) )
    
    mean_vals[, col] <- round(means$mean_value, 2)
  }

      
  # Peak CHLa (90%-tile) by year
    percentile_90 <- 
      df.james %>%
      group_by(year(date)) %>%
      summarize(CHLa_90th = round(quantile(CHLa, probs = .90, na.rm = TRUE), 2) )
  

    
summary_stats <- mean_vals %>%
  left_join(percentile_90, by = join_by("Year" == "year(date)"))

print(head(summary_stats))

```

```{r `table: Summary stats`, include = TRUE}

# Add superscripts to column names to correspond to footnotes specififying units 
    ## discharge
    colnames(summary_stats)[2] <- paste0(colnames(summary_stats)[2],
                                         footnote_marker_alphabet(1) )    
    ## surface temp
    colnames(summary_stats)[4] <- paste0(colnames(summary_stats)[4],
                                         footnote_marker_alphabet(2) )
    ## CHLa
    colnames(summary_stats)[5] <- paste0(colnames(summary_stats)[5],
                                         footnote_marker_alphabet(3) )
    ## DIN
    colnames(summary_stats)[6] <- paste0(colnames(summary_stats)[6],
                                         footnote_marker_alphabet(4) )
    ## specific conductance 
    colnames(summary_stats)[7] <- paste0(colnames(summary_stats)[7],
                                         footnote_marker_alphabet(5) )

# Define footnote content. 'Regular expressions' are used to achieve special characters, like the mu or the superscript '3' for cubic meters.
  
  fn.discharge <- "m\u00B3/s"
    
  fn.temp <- "Celsius"
  
  fn.chla <- "\u03BCg/L"
  
  fn.din <- "mg/L"
  
  fn.spcond <- "\u03BCS/cm"
  

# Print `summary_stats` dataframe as a kable table
summary_stats %>%
  kable(align = "c", 
        caption = "James River Station JMS75 Mean Water Quality Values, by Year <br> 
        Means based on values observed during July-October index period",
        escape = FALSE,
        padding = 3) %>% # 'escape' argument necessary for <sup></sup> markup to render the superscripts properly in the column names
  kable_paper() %>%
  # This next section sets what appears in the footnotes using the variables defined above. 
  footnote(alphabet = c(fn.discharge,
                        fn.temp, 
                        fn.chla, 
                        fn.din, 
                        fn.spcond)) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed") )


```

```{r `derive bloom, N-dep, and low flow counts`}

bloom_days <- 
  df.james %>%
  filter(CHLa > 40 & month(date) %in% c(7:10) ) %>%
  group_by(year(date) ) %>%
  summarize(count = n() ) %>%
  mutate("Bloom days" = count*7) %>% # readings are weekly, so multiply count by 7 to estimate totals
  rename(Year = "year(date)") %>%
  select(-count)


n_depleted <- 
  df.james %>%
  filter(DIN < .07 & month(date) %in% c(7:10) ) %>%
  group_by(year(date) ) %>%
  summarize(count = n() ) %>%
  mutate("N-depleted days" = count*7) %>%
  rename(Year = "year(date)") %>%
  select(-count)


low_flow <- 
  data.frame(Year = years_vec,
             "Low flow days" = NA)

    for (i in 2:15) {
      low_flow[i-1, 2] <- 
        length(
          which(df.discharge[jul1_yday:oct31_yday, i] < 100 ) )
    }
      

joined.df <- bloom_days %>%
  left_join(n_depleted, by = join_by("Year")) %>%
  left_join(low_flow, by = join_by("Year"))

```

<br><br>

```{r `table: bloom, N-dep, and low flow`, include = TRUE}
# Add superscripts to col names for unit specification
    ## bloom days
    colnames(joined.df)[2] <- paste0(colnames(joined.df)[2], footnote_marker_alphabet(1) )
    
    ## N-depleted days
    colnames(joined.df)[3] <- paste0(colnames(joined.df)[3], footnote_marker_alphabet(2) )
    
    ## low flow days
    colnames(joined.df)[4] <- paste0(colnames(joined.df)[4], footnote_marker_alphabet(3) )

# Print table
joined.df %>%
  kable(align = "c", 
        caption = paste0("Counts of Algal Bloom, N-Depleted, and Low Flow Days, Jul-Oct 2010-", max_yr), 
        escape = FALSE,
        padding = 3) %>%
  kable_paper() %>%
  footnote(alphabet = c("CHLa > 40 \u03BCg/L", 
                        "DIN < .07 mg/L", 
                        "discharge < 100 m\u00B3/s") ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed") )

```

<br><br>

## Model 1: CHLa \~ Temperature/Discharge Ratio

```{r `derive T:Q _model1`}
# Data wrangling for Model 1
## Subset date, SurfaceTemp, CHLa 
df.model1 <- 
  df.james[, c("date", "SurfaceTemp", "CHLa", "ln_chla", "Discharge", "period")]

## Derive T/Q ratio and ln(CHLa)
df.model1 <-
   df.model1 %>%
    mutate(t_q = SurfaceTemp/Discharge)
```

<br>

#### Regression Plots

```{r `derive regression stats1_model1`}
# Create dataframes to hold regression statistics

### For individual yr stats
 reg.stats <- 
  data.frame(Model.Year =  years_vec,
             coefficient = as.numeric(NA),
             coef.error = as.numeric(NA),
             p.val = as.numeric(NA),
             RSQR = as.numeric(NA)
             )
### For all years combined
reg.stats2 <- 
  data.frame(Model.Year =  "All years",
             coefficient = as.numeric(NA),
             coef.error = as.numeric(NA),
             p.val = as.numeric(NA),
             RSQR = as.numeric(NA)
             )



# custom functions (see 'functions.R' file)
reg.stats <- regstat_eachyear(reg.df = reg.stats, 
                              data = df.model1, 
                              x = "t_q") 

reg.stats2 <- regstat_combined(reg.df = reg.stats2, 
                               data = df.model1, 
                               x = "t_q")


# Bind together to create a single df 
reg.stats <- rbind(reg.stats2, reg.stats)

```

<br>

```{r `plot CHLa~T:Q regression_model1`, include = TRUE}
# Plot of ln(CHLa) ~ T:Q for all years' data with best fit regression line. 
# custom function (see 'functions.R' file)

  plot_allYears(  
    data = df.model1,
    x = "t_q",
    y = "ln_chla",
    formula = y ~ log(x),
    method = "lm",
    x.title = "Temperature:Discharge",
    y.title = "ln(CHLa) (\u03BCg/L)",
    axis.title.size = 14,
    point.col = c("#ED254E", "#0096FF"))

ggsave(filename = "output/model1_regressionPlot.jpeg", device = "jpeg", dpi = 300)
  
```

<br>

```{r `regression plots by year_model1`, include = TRUE, fig.width=4, fig.height=4}
#### **Chlorophyll A ~ Temperature/Discharge Ratio Plots by Year**
plot_eachYear(data = df.model1, 
                 x = "t_q", 
                 formula = y ~ log(x), 
                 method = "lm",
                 reg.stats_ = reg.stats, 
                 xlab = "Temperature:Discharge")

```

<br><br>

#### Regression Statistics

```{r `table: Regression stats_model1`, include = TRUE}
# View in table
reg.stats %>%
  kable(align = "c", 
        caption = "Regression statistics by year for ln(CHLAa) ~ ln(T:Q)") %>%
  kable_paper() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")
  )

```

<br><br>

#### Model Performance and Residuals Plots

```{r `model residuals and predicted values_model1`}

model1_tq <- lm(ln_chla ~ log(t_q), data = df.model1)
residuals_tq <- model1_tq$residuals
predicted_vals_tq <- fitted(model1_tq)

```

<br>

```{r `plot & table: bloom day prediction_model1`, fig.width=10.5, fig.height=5.5, include = TRUE, warning = FALSE}
# CHLa TIME SERIES & MODEL ACCURACY PLOT
# custom function (see 'functions.R' file)
 plot_bloomday_predictions(data = df.model1,
                           x = "t_q",
                           method = "lm") 


```

<br><br>

```{r `plot: residuals time series model1, fig.width=10.5, fig.height=4.5, warning=FALSE, include = TRUE}
# Create df with residuals and dates
df.residuals_tq <- as.data.frame(residuals_tq)
colnames(df.residuals_tq)[1] <- "residuals"
df.residuals_tq$date <- df.model1$date[as.numeric(rownames(df.residuals_tq))]

# custom function
plot.residuals_timeSeries(data = df.residuals_tq,
                          title = NULL)


```

<br><br>

```{r box-whisker plot_model1, fig.width=10.5, fig.height=5.5, warning = FALSE, include = TRUE}

# -- This requires associating each residual with a specific date/year. 
  ## -- Determine the date for each observation used in the T:Q model
    indices <- which(!is.na(df.model1$ln_chla) & !is.na(df.model1$t_q))
    tq_dates <- as.POSIXct(df.model1$date[indices])
  ## -- Join dates to residuals vector
    residuals_tq_date <- data.frame(residuals_tq, tq_dates)
    jul_oct_Residuals <- residuals_tq_date[month(residuals_tq_date$tq_dates) %in% c(7:10), ]
    
    

# Create a box-whisker plot
ggplot(jul_oct_Residuals, aes(x = format(tq_dates, "%Y"), y = residuals_tq)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "solid", color = "salmon") +  # Add a horizontal line at y=0
  theme_classic() +
  theme(legend.position = "none",
        plot.title = element_text(size = 16),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 11.5),
        axis.text.y = element_text(size = 11.5),
        axis.line = element_line(color = "grey"),
        axis.title.x = element_text(size = 16, color = "grey20", margin = margin(t = 15)),
        axis.title.y = element_text(size = 16, color = "grey20", margin = margin(r = 15))) +
  labs(title = "CHLa ~ T:Q Box-and-Whisker Index Period Residuals, ",
       x = "Year",
       y = "Residuals")

```

<br><br>

## Model 2: CHLa \~ Temperature/Discharge 4day Moving Avg. Ratio

```{r `data preparation_model2`}
## Subset date, SurfaceTemp, CHLa 
df.model2 <- 
  df.james[, c("date", "SurfaceTemp", "CHLa", "ln_chla", "Discharge", "period")]

# Create field containing DOY for each observation in df.james
df.model2 <- df.model2 %>%
  mutate(yday = yday(date))


# Calculate 4 day rolling avg for discharge 
for (i in years_vec) {
  # subset DOYs by year and assign to variable
  DOY_vec <- df.model2$yday[year(df.model2$date) == i]
  
  # Calculate rolling average (`width` argument sets window width, i.e. number of days to calcualte average )
  d <- 
    zoo::rollapply(df.discharge[, as.character(i)], width = 4, FUN = mean, align = "right", fill = NA)

  # Add discharge values to data set with temperature and chla values
  df.model2$DischargeAvg[year(df.model2$date) == i] <- d[DOY_vec]
}


# Derive T/Qavg ratio and ln(CHLa)
df.model2 <-
  df.model2 %>%
    mutate(t_q_avg = SurfaceTemp/DischargeAvg) %>%
    mutate(ln_chla = log(CHLa)) %>%
    select(-yday)

# Create factor for use in visually distinguishing 2010-2018 plot points from those 2019 and beyond.
df.model2$period <- NA

for (i in seq_along(df.model2$date)) {
  df.model2$period[i] <- ifelse(
    year(df.model2$date[i]) %in% c(2010:2018), 
    "2010-2018", 
    paste0("2019-", as.character(max_yr)))
}

```

<br>

#### Regression Plots

<br>

```{r `derive regression stats_model2}
# Create dataframes to hold regression statistics
### For individual yr stats
 reg.stats <- 
  data.frame(Model.Year =  years_vec,
             coefficient = as.numeric(NA),
             coef.error = as.numeric(NA),
             p.val = as.numeric(NA),
             RSQR = as.numeric(NA)
             )
# All years df
reg.stats2 <- 
  data.frame(Model.Year =  "All years",
             coefficient = as.numeric(NA),
             coef.error = as.numeric(NA),
             p.val = as.numeric(NA),
             RSQR = as.numeric(NA)
             )

# custom functions (see 'functions.R' file)

reg.stats <- regstat_eachyear(reg.df = reg.stats, 
                              data = df.model2, 
                              x = "t_q_avg")
reg.stats2 <- regstat_combined(reg.df = reg.stats2, 
                               data = df.model2, 
                               x = "t_q_avg")

reg.stats <- rbind(reg.stats2, reg.stats)

```

```{r `plot: regression plot all yrs combined_model2`, include = TRUE}
# Plot for ln(CHLa) ~ T:Qavg, all years with best fit regression line
# custom function (see 'functions.R' file)

  plot_allYears(  
    data = df.model2,
    x = "t_q_avg",
    y = "ln_chla",
    formula = y ~ log(x),
    method = "lm",
    title = "Chlorophyll A ~ Temperature:Qavg",
    x.title = expression("Temperature:Q"[avg]),
    y.title = "ln(CHLa) (\u03BCg/L)",
    point.col = c("#ED254E", "#0096FF"))


```

<br><br>

```{r `plots: individual years_model2`, fig.width=4, fig.height=4, include = TRUE}
#### Chlorophyll A ~ Temperature/Discharge Ratio Plots by Year
# custom function (see 'functions.R' file)

plot_eachYear(data = df.model2, 
                 x = "t_q_avg", 
                 formula = y ~ log(x), 
                 method = "lm",
                 reg.stats_ = reg.stats, 
                 xlab = expression("Temperature:Q"[avg])
              )

```

<br><br>

#### Regression Statistics

```{r `table: regression stats_model2, include = TRUE}

reg.stats %>%
  kable(align = "c", 
        caption = "Regression statistics by year for ln(CHLAa) ~ ln(T/Qavg)") %>%
  kable_paper() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")
  )

```

<br><br>

#### Model Performance and Residuals Time Series

```{r `model residuals and predicted values_model2`}

# Create a factor, "bloom_days", to indicate model prediction vs observed value
model_tq_avg <- lm(ln_chla ~ log(t_q_avg), data = df.model2)
residuals_tq_avg <- model_tq_avg$residuals
predicted_values_tq_avg <- fitted(model_tq_avg)

```

```{r `plot & table: Bloom day prediction_model1`, fig.width=10.5, fig.height=5.5, include = TRUE, warning = FALSE}
# CHLa TIME SERIES & MODEL ACCURACY PLOT.
# custom function (see 'functions.R' file)
 plot_bloomday_predictions(data = df.model2, 
                           x = "t_q_avg",
                           method = "lm") 
```

<br><br>

```{r `plot: residuals time series_model2`, fig.width=10.5, fig.height=4.5, include = TRUE, warning=FALSE}

# Create df with residuals and dates
df.residuals_tq_avg <- as.data.frame(residuals_tq_avg)
colnames(df.residuals_tq_avg)[1] <- "residuals"
df.residuals_tq_avg$date <- df.model2$date[as.numeric(rownames(df.residuals_tq_avg))]


# custom function (see 'functions.R' file)
plot.residuals_timeSeries(data = df.residuals_tq_avg,
                          title = NULL)
```

<br><br>

## Model 3: CHLa \~ Temperature/Inverse Special Conductivty Ratio

```{r `data preparation_model3`}
## Subset date, SurfaceTemp, CHLa 
df.model3 <- df.james[, c("date", "SurfaceTemp", "CHLa", "TtoCond")]

# Derive ln(CHLa)
df.model3$ln_chla <- log(df.model3$CHLa)

# Create factor for use in visually distinguishing 2010-2018 plot points from those 2019 and beyond.
df.model3$period <- NA

for (i in seq_along(df.model3$date)) {
  df.model3$period[i] <- ifelse(
    year(df.model3$date[i]) %in% c(2010:2018), 
    "2010-2018", 
    paste0("2019-", as.character(max_yr)))
}

```

<br>

#### Regression Plots

```{r `derive Regression Stat_model3}
# Create dfs to contain regression stats
 reg.stats <- 
  data.frame(Model.Year =  years_vec,
             coefficient = as.numeric(NA),
             coef.error = as.numeric(NA),
             p.val = as.numeric(NA),
             RSQR = as.numeric(NA)
             )

reg.stats2 <- 
  data.frame(Model.Year =  "All years",
             coefficient = as.numeric(NA),
             coef.error = as.numeric(NA),
             p.val = as.numeric(NA),
             RSQR = as.numeric(NA)
             )

# Custom functions to generate stats
reg.stats <- regstat_eachyear(reg.stats, data = df.model3, x = "TtoCond")
reg.stats2 <- regstat_combined(reg.stats2, data = df.model3, x = "TtoCond")


# Bind together 'all years' and individual year regression statistics 
reg.stats <- rbind(reg.stats2, reg.stats)

```

<br>

```{r `plot: regression plot all years combined_model3`, include = TRUE}
# Plot for ln(CHLa) ~ TtoCond, all years with best fit regression line.
# custom function (see 'functions.R' file)

  plot_allYears(  
    data = df.model3,
    x = "TtoCond",
    y = "ln_chla",
    formula = y ~ log(x),
    method = "lm",
    title = "CHLa~ Temperature/Inverse Special Conductivity Ratio",
    x.title = "Temperature:InvSpCond",
    y.title = "ln(CHLa) (\u03BCg/L)",
    point.col = c("#ED254E", "#0096FF"))

```

<br>

```{r `plots: individual years`, fig.width=4, fig.height=4, include = TRUE}
#### Chlorophyll A ~ Temperature/Inverse Special Conductivity Ratio, by Year
# custom function (see 'functions.R' file)

plot_eachYear(data = df.model3, 
                 x = "TtoCond", 
                 formula = y ~ log(x), 
                 method = "lm",
                 reg.stats_ = reg.stats, 
                 xlab = "Temperature:InvSpCond")

```

<br><br>

#### Regression Statistics

```{r `table: regression stats_model3`, include = TRUE}
  
# View in table
reg.stats %>%
  kable(align = "c", 
        caption = "Regression statistics by year for ln(CHLAa) ~ ln(T:InvSpCond)") %>%
  kable_paper() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")
  )

```

<br><br>

#### Model Performance and Residuals Time Series

```{r `model residuals and predicted values_model3`}
# Create a factor, "bloom_days", to indicate model prediction vs observed value
model_tcond <- lm(ln_chla ~ log(TtoCond), data = df.model3)
residuals_tcond <- model_tcond$residuals
predicted_values_tcond <- fitted(model_tcond)

```

```{r `plot & table: bloom day prediction_model3`, fig.width=10.5, fig.height=5.5, include = TRUE, warning = FALSE}
# CHLa TIME SERIES & MODEL ACCURACY PLOT.
# custom function (see 'functions.R' file)
 plot_bloomday_predictions(data = df.model3, 
                           x = "TtoCond",
                           method = "lm")   # Custom function
```

<br><br>

```{r `plot: residuals time series_model3, fig.width=10.5, fig.height=4.5, include = TRUE, warning=FALSE}
# Create df with residuals and dates
df.residuals_tcond <- as.data.frame(residuals_tcond)
colnames(df.residuals_tcond)[1] <- "residuals"
df.residuals_tcond$date <- df.model3$date[as.numeric(rownames(df.residuals_tcond))]

# custom function (see 'functions.R' file)
plot.residuals_timeSeries(data = df.residuals_tcond, 
                          title = NULL) 

```

<br><br>

## Model 4: Nonlinear Tangential Model

<br>

#### $$CHLa \sim CHLamax \times \tanh\left(\frac{\alpha \times T}{CHLamax}\right)$$

```{r `modeling and predictions_model4`}
# Recycle df.model1 dataframe, as it already has t_q values.
df_tang <- df.model1[, -9]

model_tang <- nls(CHLa ~ CHLamax * tanh((alpha * t_q)/CHLamax), 
             data = df_tang, 
             start = c(alpha = 50, CHLamax = 10),
             control = nls.control(maxiter = 2000))

mod_tang_sum <- summary(model_tang)


mod_tang_predictions <- fitted(model_tang)


obs_fit <- data.frame(observed = df_tang$CHLa[!is.na(df_tang$CHLa)], 
                      predicted = fitted(model_tang),
                      date = df_tang$date[!is.na(df_tang$CHLa)])
```

<br>

```{r `scatterplots w/ model fit_model4`, include = TRUE}
# Plot for ln(CHLa) ~ T:Qavg, all years
 plot_allYears <- ggplot(df_tang, aes(x = t_q, y = CHLa)) +
  geom_point(aes(col=period), alpha = 0.55) +
  geom_smooth(method = "nls", formula = y ~ CHLamax * tanh((alpha * x)/CHLamax),
              se = FALSE, linetype = "solid", color = "grey45", alpha = 0.5,
              method.args = list(start = coef(model_tang))) +
  labs(title = NULL,
           x = "Temperature:Discharge",
           y = "CHLa (\u03BCg/L)") +
  scale_color_manual(values = c("#ED254E", "#0096FF"), 
                         name = NULL) +
  theme(title = element_text(size = 8)) +
  theme_minimal()

suppressWarnings(
  print(plot_allYears)
  )
```

<br>

```{r `derive regression stats_model4`}
## LINEAR REGRESSION STATISTICS 
 reg.stats_tang <- 
  data.frame(Model.Year =  years_vec,
             alpha.coef = as.numeric(NA),
             alpha.p.val = as.numeric(NA),
             alpha.se = as.numeric(NA),
             CHLamax.coef = as.numeric(NA),
             CHLamax.pval = as.numeric(NA),
             CHLamax.se = as.numeric(NA) )

# Run the model for each year
for (i in seq_along(years_vec)) {
  data_subset <- 
    df_tang[year(df_tang$date) == years_vec[i], ]
  
  res_tang <- 
    summary(
      nls(CHLa ~ CHLamax * tanh((alpha * t_q)/CHLamax), 
          data = data_subset, 
          start = c(alpha = 70, CHLamax = 20),
          control = nls.control(maxiter = 1000) )
    )
  
  reg.stats_tang$alpha.coef[i] <- round(res_tang$coefficients[1,1], 4)
   reg.stats_tang$alpha.p.val[i] <- round(res_tang$coefficients[1,4], 4)
    reg.stats_tang$alpha.se[i] <- round(res_tang$coefficients[1,2] ,4)
     reg.stats_tang$CHLamax.coef[i] <- round(res_tang$coefficients[2,1], 4)
      reg.stats_tang$CHLamax.pval[i] <- round(res_tang$coefficients[2,4], 4)
       reg.stats_tang$CHLamax.se[i] <- round(res_tang$coefficients[2,2], 4)

}
  
# Add regression stats for all years -- Repeat same as above without for loop and create a 1 row data frame.
reg.stats_tang2 <- 
  data.frame(Model.Year =  "All years",
             alpha.coef = as.numeric(NA),
             alpha.p.val = as.numeric(NA),
             alpha.se = as.numeric(NA),
             CHLamax.coef = as.numeric(NA),
             CHLamax.pval = as.numeric(NA),
             CHLamax.se = as.numeric(NA) )

 res_tang <- 
    summary(
      nls(CHLa ~ CHLamax * tanh((alpha * t_q)/CHLamax), 
          data = df_tang, 
          start = c(alpha = 70, CHLamax = 10),
          control = nls.control(maxiter = 1000) )
    )
  
  
  reg.stats_tang2$alpha.coef <- round(res_tang$coefficients[1,1], 4)
   reg.stats_tang2$alpha.p.val <- round(res_tang$coefficients[1,4], 4)
    reg.stats_tang2$alpha.se <- round(res_tang$coefficients[1,2] ,4)
     reg.stats_tang2$CHLamax.coef <- round(res_tang$coefficients[2,1], 4)
      reg.stats_tang2$CHLamax.pval <- round(res_tang$coefficients[2,4], 4)
       reg.stats_tang2$CHLamax.se <- round(res_tang$coefficients[2,2], 4)

# Bind together 'all years' and individual year regression statistics 
reg.stats_tang <- rbind(reg.stats_tang2, reg.stats_tang)

```

```{r `plots: individual years_model4`, fig.width=4, fig.height=4, include = TRUE}

for (yr in years_vec) {
  plot <- 
    ggplot(df_tang, aes(x = t_q, y = CHLa)) +
      geom_smooth(data = df_tang[year(df_tang$date) == yr,],
                  method = "nls", 
                  formula = y ~ CHLamax * tanh((alpha * x)/CHLamax),
                  se = FALSE, linetype = "solid", color = "grey45", alpha = 0.75,
                  method.args = list(start = coef(model_tang))) +
      geom_point(data = df_tang[year(df_tang$date) == yr,],
                 color = ifelse(yr < 2019, "#ED254E", "#0096FF"), 
                 alpha = .6) +
      labs(title = yr,
           x = "Temperature:Discharge",
           y = "CHLa (\u03BCg/L)") +
      theme(title = element_text(size = 8)) +
      theme_minimal()
  
  suppressWarnings(
    print(plot)
  )
}


```

<br>

```{r `residuals vs predicted_model4`, include = TRUE}
ggplot(obs_fit, aes(x = predicted, y = residuals(model_tang) ) ) +
  geom_point(alpha = .65) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Predicted Values", y = "Residuals", title = "Residuals vs. Predicted") + 
  theme_classic()


```

<br><br>

#### Parameter Estimates

```{r `table: model parameter estimates_model4`, include = TRUE}
reg.stats_tang %>%
  kable(align = "c", 
        caption = "CHLa ~ CHLamax * tanh((alpha * t_q)/CHLamax) Model Statistics, by Year") %>%
  kable_paper() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")
  )

```

<br><br>

#### Model Performance and Residuals Time Series

<br>

```{r `plot: residuals time series_model4, fig.width=10.5, fig.height=4.5, include = TRUE, warning=FALSE}

# Create df with residuals and dates
df.residuals_tang <- as.data.frame(residuals(model_tang))
colnames(df.residuals_tang)[1] <- "residuals"
df.residuals_tang$date <- df_tang$date[as.numeric(rownames(df.residuals_tang))]

# custom function (see 'functions.R' file)
plot.residuals_timeSeries(data = df.residuals_tang,
                          title = NULL)

```

<br>

```{r `plot & table: bloom day prediction_model4`, fig.width=10.5, fig.height=5.5, include = TRUE, warning = FALSE}
## TIME SERIES PLOT & MODEL ACCURACY TABLE
# custom function (see 'functions.R' file)
plot_bloomday_predictions(data = df_tang,
                          x = "t_q",
                          method = "nls")

```
